---
title: "Getting started with composableIDModelR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with composableIDModelR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4
)
```

composableIDModelR builds infectious disease models from interchangeable parts and runs
them with
[ComposableTuringIDModels.jl](https://github.com/EpiAware/ComposableTuringIDModels.jl)
in Julia.
A model has two parts:

- an **infection model**, which generates unobserved infections and owns the
  **latent process** that drives them (for example the log reproduction number
  of a renewal model), and
- an **observation model**, which links infections to data, possibly through
  delays or under-ascertainment.

Every part has an R constructor with the same name and arguments as its Julia
counterpart, so the
[ComposableTuringIDModels.jl documentation](https://composableturingidmodels.epiaware.org/stable/)
applies directly.

```{r load}
library(composableIDModelR)
```

## Setting up Julia

composableIDModelR needs Julia (>= 1.12), the version its pinned dependencies were
resolved with, ideally installed with
[juliaup](https://github.com/JuliaLang/juliaup).
The Julia packages are installed and loaded the first time a model is simulated
or fitted, which can take several minutes.
To follow progress, run the setup yourself:

```{r julia_setup, eval = FALSE}
cidm_setup_julia()
```

To run MCMC chains in parallel, set `JULIA_NUM_THREADS` before Julia starts:

```{r threads, eval = FALSE}
Sys.setenv(JULIA_NUM_THREADS = "4")
```

## Building a model

Distributions come from [distspec](https://epiforecasts.io/distspec/), so
they can be given by their natural parameters or by a mean and standard
deviation.

```{r components}
latent <- RandomWalk(
  init = Normal(0, 0.25),
  epsilon_t = HierarchicalNormal(std = HalfNormal(0.05))
)

infection <- Renewal(
  generation_time = Gamma(shape = 6.5, scale = 0.62),
  rt = latent,
  initialisation = Normal(log(10), 0.5)
)

observation <- NegativeBinomialError(cluster_factor = HalfNormal(0.1))

model <- IDModel(infection, observation)
model
```

Building a model does not start Julia.
The printout is the Julia code the model runs, which you can also get as a
string with `as_julia()`.
Arguments left out use the Julia defaults.

## Simulating from the prior

`simulate()` draws parameters from their priors and simulates the latent
process, infections and observations.
Each quantity is returned as a matrix with one row per simulation.

```{r simulate}
sims <- simulate(model, nsim = 20, n = 50, seed = 1)
names(sims)
matplot(t(sims$Rt), type = "l", lty = 1, col = "grey50", log = "y",
        xlab = "Day", ylab = "Rt")
```

Prior simulations are a quick check that the priors produce plausible
epidemics before fitting.

## Fitting to data

We fit the model to one of the simulated series.
`nuts()` sets up the sampler; short chains keep this example fast.

```{r fit}
y <- sims$generated_y_t[1, ]

fitted <- fit(
  model, y,
  method = nuts(draws = 250, warmup = 250, chains = 2),
  seed = 2
)
fitted
```

`summary()` gives posterior summaries of the parameters, and the draws work
with the [posterior](https://mc-stan.org/posterior/) package and anything
built on it, such as bayesplot.

```{r summary}
summary(fitted)
```

Generated quantities (the latent process `Z_t`, infections `I_t`, expected and
predicted observations, and `Rt` for renewal models) are matrices with one row
per posterior draw:

```{r generated}
str(fitted$generated)
```

## Plotting and forecasting

```{r plot_rt}
plot(fitted, type = "Rt")
```

`type = "cases"` compares posterior predictions with the data, and `horizon`
adds a forecast, which continues the latent process beyond the data.

```{r plot_cases}
plot(fitted, type = "cases", horizon = 14)
```

The forecast draws themselves come from `predict()`:

```{r predict}
forecasts <- predict(fitted, horizon = 14)
dim(forecasts)
```

## Swapping components

Changing an assumption means replacing one part.
Here observations follow a delay from infection and are Poisson distributed,
and the reproduction number follows an AR(1) process:

```{r swap}
IDModel(
  Renewal(
    generation_time = Gamma(shape = 6.5, scale = 0.62),
    rt = AR(damp = truncated(Normal(0.9, 0.05), 0, 1)),
    initialisation = Normal(log(10), 0.5)
  ),
  LatentDelay(PoissonError(), delay = LogNormal(1.6, 0.42))
)
```

## Components without an R wrapper

`component()` constructs any Julia component by name, and `julia()` inserts
Julia code verbatim, for example a function argument.
Positional R arguments become positional Julia arguments and named ones become
keyword arguments.

```{r escape_hatch}
IDModel(
  DirectInfections(
    Z = component("HilbertSpaceGP", role = "latent"),
    initialisation = Normal(log(100), 1)
  ),
  LatentDelay(PoissonError(), delay = component("Frechet", 3, 2,
                                                role = "distribution"))
)
```

The [Mishra et al. (2020) case study](mishra-case-study.html) applies these
steps to COVID-19 data from South Korea.
