---
title: "Replicating EpiNow2 Models"
author: "Sebastian Funk"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Replicating EpiNow2 Models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

## Introduction

[EpiNow2](https://epiforecasts.io/EpiNow2/) is a popular R package for estimating time-varying reproduction numbers using renewal equation models. This vignette shows how to replicate a typical EpiNow2 analysis using EpiAwareR's compositional approach.

Both packages use renewal equations and Bayesian inference, but EpiAwareR's compositional design allows you to build a much wider class of models by assembling components.

```{r load, message = FALSE}
library(EpiAwareR)
```

## Example: A Typical EpiNow2 Analysis

We'll replicate a standard EpiNow2 workflow for estimating $R_t$ from case data.

```{r data}
set.seed(123)
dates <- seq.Date(as.Date("2024-01-01"), by = "day", length.out = 50)

# Epidemic with intervention at day 25
infections <- numeric(50)
infections[1:7] <- 30
for (i in 8:50) {
  rt <- if (i < 25) 2.0 else 0.7 # Intervention effect
  infections[i] <- rpois(1, rt * mean(infections[max(1, i - 7):(i - 1)]))
}

outbreak_data <- data.frame(
  date = dates,
  confirm = as.integer(pmax(round(infections + rnorm(50, 0, 5)), 1)) # Integer counts
)

head(outbreak_data, 10)
```

### Step 1: The EpiNow2 Approach

A typical EpiNow2 analysis with a daily random walk looks like this:

```{r epinow2_check, include = FALSE}
has_epinow2 <- requireNamespace("EpiNow2", quietly = TRUE)
```

```{r epinow2_example, eval = has_epinow2}
library(EpiNow2)

# Define generation time
generation_time <- Gamma(
  mean = 5.0,
  sd = 2.0
)

# Define reporting delay
reporting_delay <- LogNormal(
  mean = 2.0,
  sd = 1.0
)

# Use daily random walk for Rt (alternative to GP)
rt_settings <- rt_opts(
  prior = Normal(mean = 1, sd = 1),
  rw = 1 # Daily random walk
)

# Run estimation with timing
epinow2_time <- system.time({
  epinow2_results <- epinow(
    outbreak_data,
    generation_time = generation_time_opts(generation_time),
    delays = delay_opts(reporting_delay),
    rt = rt_settings,
    stan = stan_opts(cores = 4, chains = 4)
  )
})

cat("EpiNow2 runtime:", epinow2_time["elapsed"], "seconds\n")

# View results
summary(epinow2_results)
plot(epinow2_results)
```

The `rw = 1` option creates a daily random walk for $R_t$, which is an alternative choice to the default Gaussian Process.

### Step 2: The EpiAwareR Equivalent

To replicate this in EpiAwareR, we explicitly build the model from components:

```{r epiaware_equivalent}
# 1. Latent process: daily random walk for Rt
# AR(1) with damping ≈ 1 creates a random walk (equivalent to EpiNow2's rw=1)
latent <- AR(
  order = 1,
  damp_priors = list(truncnorm(0.99, 0.01, 0.9, 1)), # Near 1 for random walk
  init_priors = list(norm(0, 1)), # log(Rt) prior: Rt centered at 1
  std_prior = halfnorm(0.1)
)

# 2. Infection model: renewal equation with generation time
# EpiNow2: Gamma(mean=5, sd=2) -> shape=6.25, scale=0.8
infection <- Renewal(
  gen_distribution = gamma_dist(6.25, 0.8),
  initialisation_prior = norm(log(100), 2)
)

# 3. Observation model: negative binomial with reporting delay
# EpiNow2: LogNormal(mean=2, sd=1) -> meanlog=0.58, sdlog=0.47
observation <- LatentDelay(
  model = NegativeBinomialError(
    cluster_factor_prior = halfnorm(0.5)
  ),
  delay_distribution = lognorm(0.58, 0.47)
)

# 4. Compose into complete model
model <- EpiProblem(
  epi_model = infection,
  latent_model = latent,
  observation_model = observation,
  tspan = c(1, 50)
)

print(model)
```

### Fit the Model

```{r fit}
# Run estimation with timing (4 chains to match EpiNow2)
epiaware_time <- system.time({
  results <- fit(
    model = model,
    data = outbreak_data,
    method = nuts_sampler(
      warmup = 500,
      draws = 500,
      chains = 4
    )
  )
})

cat("EpiAwareR runtime:", epiaware_time["elapsed"], "seconds\n")

# View results
print(results)
summary(results)
plot(results, type = "Rt")
```

### Runtime Comparison

```{r runtime_comparison, eval = has_epinow2}
cat("Runtime comparison (4 chains each):\n")
cat("  EpiNow2:   ", round(epinow2_time["elapsed"], 1), "seconds\n")
cat("  EpiAwareR: ", round(epiaware_time["elapsed"], 1), "seconds\n")
cat("  Speedup:   ", round(epinow2_time["elapsed"] / epiaware_time["elapsed"], 1), "x\n")
```

## Benefits of the explicit approach

EpiAwareR's compositional design makes the model structure transparent:

1. **Latent model** explicitly states how $R_t$ evolves (random walk, AR, MA, etc.)
2. **Infection model** clearly specifies generation time and renewal dynamics
3. **Observation model** shows delays and overdispersion assumptions

This explicitness enables:

- **Testing alternatives**: Easily compare different smoothness levels, or swap e.g. for AR(2), moving average or mechanistic models (e.g. with susceptible depletion)
- **Understanding priors**: See exactly what distributions drive each component
- **Model comparison**: Systematically compare competing assumptions

For example, comparing different levels of smoothness in the random walk:

```{r rw_comparison, eval = FALSE}
# Less smooth (more variable Rt)
rw_flexible <- AR(
  order = 1,
  damp_priors = list(truncnorm(1.0, 0.01, 0.99, 1)),
  init_priors = list(norm(0, 1)),
  std_prior = halfnorm(0.3) # Larger innovations
)

# More smooth (less variable Rt)
rw_smooth <- AR(
  order = 1,
  damp_priors = list(truncnorm(1.0, 0.01, 0.99, 1)),
  init_priors = list(norm(0, 1)),
  std_prior = halfnorm(0.1) # Smaller innovations
)

# Fit both and compare
model_flexible <- EpiProblem(..., latent_model = rw_flexible, ...)
model_smooth <- EpiProblem(..., latent_model = rw_smooth, ...)

results_flexible <- fit(model_flexible, outbreak_data)
results_smooth <- fit(model_smooth, outbreak_data)

# Visual inspection of results
# Formal model comparison would require LOO cross-validation (for fit)
# or leave-future-out validation (for forecasting)
print(results_flexible)
print(results_smooth)
```

## Learn More

- **[EpiNow2 Documentation](https://epiforecasts.io/EpiNow2/)** - The original package
- **[Getting Started](getting-started.html)** - EpiAwareR basics
- **[Mishra Case Study](mishra-case-study.html)** - Detailed worked example

## Session Info

```{r session_info}
sessionInfo()
```
