---
title: "Case study: Mishra et al. (2020)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Case study: Mishra et al. (2020)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4
)
```

Mishra et al. (2020) derived the renewal equation from an age-dependent
branching process and combined it with a negative binomial observation model to
estimate the time-varying reproduction number $R_t$ from reported cases.
This vignette fits their model to test-confirmed COVID-19 cases from South Korea
in early 2020, following the
[ComposableTuringIDModels.jl tutorial](https://composableturingidmodels.epiaware.org/stable/tutorials/renewal-negbin)
of the same analysis.

```{r load, message = FALSE}
library(composableIDModelR)
```

## Data

The package includes daily confirmed cases in South Korea.
We fit the first wave, from 13 February to 19 March 2020.

```{r data}
south_korea <- read.csv(
  system.file("extdata", "south_korea_data.csv", package = "composableIDModelR")
)
south_korea$date <- as.Date(south_korea$date)
first_wave <- south_korea[45:80, ]

plot(first_wave$date, first_wave$cases_new, type = "b",
     xlab = "Date", ylab = "Confirmed cases")
```

## Model

The model is

$$
\begin{aligned}
Z_t &= \rho_1 Z_{t-1} + \rho_2 Z_{t-2} + \epsilon_t,
  & \epsilon_t &\sim \mathrm{Normal}(0, \sigma), \\
R_t &= \exp(Z_t), \\
I_t &= R_t \sum_{s \ge 1} g_s I_{t-s}, \\
y_t &\sim \mathrm{NegBinomial}(I_t, \phi),
\end{aligned}
$$

where $g_s$ is the discretised generation interval.

The latent log reproduction number follows an AR(2) process.
A first damping coefficient concentrated near one encodes strong
autocorrelation, and the innovation standard deviation $\sigma$ is estimated
with a half-normal prior.

```{r latent}
latent <- AR(
  damp = list(
    truncated(Normal(0.8, 0.05), 0, 1),
    truncated(Normal(0.1, 0.05), 0, 1)
  ),
  init = list(Normal(0, 0.2), Normal(0, 0.2)),
  epsilon_t = HierarchicalNormal(std = HalfNormal(0.1))
)
```

The latent process belongs to the renewal model.
Following Mishra et al. (2020), a gamma distribution with shape 6.5 and scale
0.62 approximates the generation interval, and is discretised with double
interval censoring.

```{r renewal}
renewal <- Renewal(
  generation_time = Gamma(shape = 6.5, scale = 0.62),
  rt = latent,
  initialisation = Normal(log(1), 0.1)
)
```

Reported cases are negative binomial.
The prior is on the cluster factor $\sqrt{1/\phi}$, which is approximately the
coefficient of variation of the observation noise.

```{r observation}
model <- IDModel(
  renewal,
  NegativeBinomialError(cluster_factor = HalfNormal(0.1))
)
model
```

## Fit

We run four chains with a raised target acceptance rate, which keeps the
sampler stable on the hierarchical innovation scale.
The chains are kept short so the vignette builds quickly.

```{r fit}
fitted <- fit(
  model,
  first_wave$cases_new,
  dates = first_wave$date,
  method = nuts(draws = 500, warmup = 500, chains = 4,
                target_acceptance = 0.95),
  seed = 1234
)
fitted
```

The printout reports some divergent transitions, so for a real analysis the
chains should be run for longer and the divergences investigated, for
example with the sampler diagnostics in `fitted$sampler`.

Parameters are named after the component that samples them: `damp_AR` are the
damping coefficients $\rho$, `std` is $\sigma$, `cluster_factor` is the
observation cluster factor, and `init_incidence` is the log initial incidence.
The standardised innovations `ϵ_t` are omitted here.

```{r summary}
summary(fitted) |>
  subset(!startsWith(variable, "ϵ_t"))
```

The draws are a `posterior::draws_df`, so all of the
[posterior](https://mc-stan.org/posterior/) tooling applies.

```{r posterior}
posterior::subset_draws(
  posterior::as_draws_df(fitted),
  variable = c("damp_AR", "std", "cluster_factor")
) |>
  posterior::summarise_draws("mean", "sd", "rhat", "ess_bulk")
```

## Results

The reproduction number rises well above one early in the wave, falls through
one as cases peak, and stays below one as cases decline.

```{r rt}
plot(fitted, type = "Rt")
```

The posterior predictive distribution tracks the observed cases.

```{r cases}
plot(fitted, type = "cases")
```

## Swapping the observation model

Reported cases lag infections by the incubation period.
Adding a log-normal delay (mean of about 5 days) changes only the observation
model:

```{r delay}
delayed_model <- IDModel(
  renewal,
  LatentDelay(
    NegativeBinomialError(cluster_factor = HalfNormal(0.1)),
    delay = LogNormal(1.6, 0.42)
  )
)

delayed <- fit(
  delayed_model,
  first_wave$cases_new,
  dates = first_wave$date,
  method = nuts(draws = 500, warmup = 500, chains = 4,
                target_acceptance = 0.95),
  seed = 1234
)
plot(delayed, type = "Rt")
```

Accounting for the delay moves the decline in $R_t$ earlier, because infections
precede the cases they cause.

## References

Mishra, S., Berah, T., Mellan, T. A., et al. (2020). On the derivation of the
renewal equation from an age-dependent branching process: an epidemic modelling
perspective. *arXiv:2006.16487*.
